---
title: "PHIPER summary report"
author: 
  - name: "Carlos S. Reyna-Blanco"
  - name: "Mateusz F. Kolek"
  - name: "Gabriel Innocenti"
date: today
date-format: "MMMM D, YYYY"
description: "Automated PhIP-Seq report for group-level comparison and feature-level interpretation."
format:
  html:
    title-block-banner: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
params:
  base_dir: null # passed at runtime, e.g. "SAR/results"
  group_col: null # output group folder name; kept for backwards compatibility
  top_n: 500
  tables_open: false
  include_beta_tables: false
  include_alpha_tables: false
  include_tsne3d: true
  include_longitudinal_stability: true
  include_pop_tables: true
  include_pop_plots: true
  include_delta_tables: true
  include_delta_feature_plots: true
  pop_interactive_mode: "link"      # one of: "none", "link", "embed"
  delta_interactive_mode: "embed"    # one of: "none", "link", "embed"
  max_delta_feature_plots: 5
  min_eff: 5
  rank_cols: null
---

```{r}
suppressPackageStartupMessages({
  library(fs)
  library(dplyr)
  library(readr)
  library(readxl)
  library(stringr)
  library(glue)
  library(knitr)
  library(htmltools)
  library(tibble)
  library(purrr)
  library(DT)
})

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0 || is.na(x)) y else x

base_dir  <- params$base_dir
group_col <- params$group_col

if (is.null(base_dir) || !nzchar(base_dir)) {
  stop("Missing required Quarto parameter: base_dir")
}

if (is.null(group_col) || !nzchar(group_col)) {
  stop("Missing required Quarto parameter: group_col")
}

group_dir <- fs::path(base_dir, group_col)

if (!fs::dir_exists(group_dir)) {
  stop(glue("Group folder not found: {group_dir}"))
}

ignored_dirs <- c("phiper_summary_report_files")
comparisons <- fs::dir_ls(group_dir, type = "directory") |>
  basename() |>
  (\(x) x[!grepl("phiper_summary", x)])() |>
  setdiff(ignored_dirs) |>
  sort()


if ("all" %in% comparisons) {
  comparisons <- c("all", setdiff(comparisons, "all"))
}

normalize_interactive_mode <- function(x, default = "link") {
  x <- tolower(as.character(x %||% default))
  if (!x %in% c("none", "link", "embed")) default else x
}


pop_interactive_mode   <- normalize_interactive_mode(params$pop_interactive_mode, "link")
delta_interactive_mode <- normalize_interactive_mode(params$delta_interactive_mode, "link")

safe_rel_path <- function(x, start = group_dir) {
  rel <- fs::path_rel(x, start = start)
  rel <- gsub("%2F", "/", utils::URLencode(rel, reserved = TRUE))
  rel
}

cat_md <- function(...) cat(..., sep = "")

emit_heading <- function(text, level = 2) {
  cat_md(strrep("#", level), " ", text, "\n\n")
}

emit_text <- function(text = "") {
  cat_md(text, "\n\n")
}

emit_missing <- function(text = "_Not available._") {
  emit_text(text)
}

emit_img <- function(file, caption = NULL, width = "100%") {
  if (length(file) == 0 || is.na(file) || !fs::file_exists(file)) {
    emit_missing()
    return(invisible(NULL))
  }
  src <- safe_rel_path(file)
  alt <- if (is.null(caption)) basename(file) else caption
  cat_md(glue('<div style="margin: 0.5rem 0 1rem 0;">\n'))
  cat_md(glue('<img src="{src}" alt="{htmlEscape(alt)}" style="width:{width}; border:1px solid #e5e7eb; border-radius:8px;"/>\n'))
  if (!is.null(caption)) {
    cat_md(glue('<div style="font-size:0.92rem;color:#555;margin-top:0.35rem;">{htmlEscape(caption)}</div>\n'))
  }
  cat_md('</div>\n\n')
}

emit_img_grid <- function(files, labels = NULL, ncol = 2) {
  files <- files[!is.na(files)]
  files <- files[fs::file_exists(files)]
  if (length(files) == 0) {
    emit_missing()
    return(invisible(NULL))
  }
  if (is.null(labels)) labels <- basename(files)
  cat_md(glue('<div style="display:grid;grid-template-columns:repeat({ncol}, minmax(0, 1fr));gap:24px;margin: 0.75rem 0 1.25rem 0;">\n'))
  for (i in seq_along(files)) {
    src <- safe_rel_path(files[[i]])
    lab <- labels[[i]]
    cat_md('<div style="border:1px solid #e5e7eb;border-radius:10px;padding:14px;">\n')
    cat_md(glue('<img src="{src}" alt="{htmlEscape(lab)}" style="width:100%;"/>\n'))
    cat_md(glue('<div style="font-size:0.95rem;color:#555;margin-top:0.5rem;">{htmlEscape(lab)}</div>\n'))
    cat_md('</div>\n')
  }
  cat_md('</div>\n\n')
}


parse_csv <- function(x, default = NULL) {
  if (is.null(x) || !nzchar(trimws(x))) return(default)

  x <- strsplit(x, ",", fixed = TRUE)[[1]]
  x <- trimws(x)
  x <- x[nzchar(x)]

  x
}

emit_iframe <- function(file, height = "700px") {
  if (length(file) == 0 || is.na(file) || !fs::file_exists(file)) {
    emit_missing()
    return(invisible(NULL))
  }
  src <- safe_rel_path(file)
  cat_md(glue('<iframe src="{src}" style="width:100%;height:{height};border:1px solid #e5e7eb;border-radius:8px;"></iframe>\n\n'))
}

emit_link_list <- function(files, title = "Open interactive plots") {
  files <- files[!is.na(files)]
  files <- files[fs::file_exists(files)]
  if (length(files) == 0) {
    emit_missing()
    return(invisible(NULL))
  }
  cat_md(glue('<details><summary><strong>{htmlEscape(title)}</strong></summary>\n'))
  cat_md('<ul>\n')
  for (f in files) {
    rel <- safe_rel_path(f)
    nm  <- basename(f)
    cat_md(glue('<li><a href="{rel}" target="_blank">{htmlEscape(nm)}</a></li>\n'))
  }
  cat_md('</ul>\n</details>\n\n')
}

read_table_auto <- function(file) {
  ext <- tolower(fs::path_ext(file))
  out <- switch(
    ext,
    csv  = readr::read_csv(file, show_col_types = FALSE, progress = FALSE),
    xlsx = readxl::read_xlsx(file),
    xls  = readxl::read_xls(file),
    rds  = readRDS(file),
    stop(glue("Unsupported table type: {file}"))
  )

  if (is.matrix(out)) out <- as.data.frame(out)
  if (inherits(out, c("dist", "table"))) out <- as.data.frame(out)

  if (!is.data.frame(out)) {
    if (is.list(out) && length(out) > 0 && is.data.frame(out[[1]])) {
      out <- out[[1]]
    } else if (is.list(out) && length(out) > 0) {
      out <- tibble::tibble(value = capture.output(str(out, max.level = 1)))
    } else {
      out <- tibble::tibble(value = capture.output(str(out)))
    }
  }

  tibble::as_tibble(out)
}

prepare_table <- function(df, top_n = params$top_n) {
  suppressWarnings({
    if ("p_raw" %in% names(df) && "p_adj_rank_wbh" %in% names(df)) {
      df <- df |> arrange(.data$p_adj_rank_wbh, .data$p_raw)
    } else if ("p_raw" %in% names(df) && "p_adj_rank" %in% names(df)) {
      df <- df |> arrange(.data$p_adj_rank, .data$p_raw)
    } else if ("p_raw" %in% names(df)) {
      df <- df |> arrange(.data$p_raw)
    } 
  })

  dplyr::slice_head(df, n = min(top_n, nrow(df)))
}


emit_dt_details <- function(file,
                            title = basename(file),
                            top_n = params$top_n,
                            filter_fun = NULL) {
  if (length(file) == 0 || is.na(file) || !fs::file_exists(file)) {
    emit_missing()
    return(invisible(NULL))
  }

  df <- read_table_auto(file)

  if (!is.null(filter_fun)) {
    df <- filter_fun(df)
  }

  df <- prepare_table(df, top_n = top_n)

  if (nrow(df) == 0) {
    emit_missing("_Table exists but is empty after filtering._")
    return(invisible(NULL))
  }

  assign(".dt_current_df", df, envir = knitr::knit_global())
  assign(".dt_current_title", title, envir = knitr::knit_global())
  assign(".dt_current_open", params$tables_open, envir = knitr::knit_global())

  child <- knitr::knit_child(
    text = c(
      "```{r, echo=FALSE, warning=FALSE, message=FALSE}",
      "if (isTRUE(.dt_current_open)) {",
      "  htmltools::tagList(",
      "    htmltools::tags$details(",
      "      open = NA,",
      "      htmltools::tags$summary(htmltools::tags$strong(.dt_current_title)),",
      "      htmltools::tags$div(",
      "        style = 'margin-top: 0.75rem;',",
      "        DT::datatable(",
      "          .dt_current_df,",
      "          rownames = FALSE,",
      "          filter = 'top',",
      "          options = list(",
      "            pageLength = 10,",
      "            lengthMenu = c(10, 25, 50, 100),",
      "            scrollX = TRUE,",
      "            autoWidth = TRUE",
      "          )",
      "        )",
      "      )",
      "    )",
      "  )",
      "} else {",
      "  htmltools::tagList(",
      "    htmltools::tags$details(",
      "      htmltools::tags$summary(htmltools::tags$strong(.dt_current_title)),",
      "      htmltools::tags$div(",
      "        style = 'margin-top: 0.75rem;',",
      "        DT::datatable(",
      "          .dt_current_df,",
      "          rownames = FALSE,",
      "          filter = 'top',",
      "          options = list(",
      "            pageLength = 10,",
      "            lengthMenu = c(10, 25, 50, 100),",
      "            scrollX = TRUE,",
      "            autoWidth = TRUE",
      "          )",
      "        )",
      "      )",
      "    )",
      "  )",
      "}",
      "```"
    ),
    quiet = TRUE
  )

  cat(child, sep = "\n")
  invisible(NULL)
}


show_table <- function(file, title = basename(file), top_n = params$top_n, max_rows = 100) {
  if (length(file) == 0 || is.na(file) || !fs::file_exists(file)) {
    emit_missing()
    return(invisible(NULL))
  }

  df <- prepare_table(read_table_auto(file), top_n = top_n)

  if (nrow(df) == 0) {
    emit_missing("_Table exists but is empty._")
    return(invisible(NULL))
  }

  if (!is.null(title) && nzchar(title)) {
    cat_md(glue('<div style="font-weight:600;margin:0.5rem 0 0.5rem 0;">{htmlEscape(title)}</div>\n'))
  }

  tbl <- knitr::kable(
    utils::head(df, max_rows),
    format = "html",
    escape = TRUE,
    table.attr = 'class="table table-striped table-hover table-sm" style="width:100%;"'
  )

  cat_md(as.character(tbl), "\n")

  if (nrow(df) > max_rows) {
    cat_md(glue(
      '<div style="font-size:0.9rem;color:#666;margin-top:0.35rem;">Showing first {max_rows} of {nrow(df)} rows.</div>\n'
    ))
  }

  cat_md("\n")
  invisible(NULL)
}

emit_table_details <- function(file, title = basename(file), top_n = params$top_n) {
  if (length(file) == 0 || is.na(file) || !fs::file_exists(file)) {
    emit_missing()
    return(invisible(NULL))
  }

  cat_md(glue('<details><summary><strong>{htmlEscape(title)}</strong></summary>\n\n'))
  show_table(file, title = NULL, top_n = top_n)
  cat_md('\n</details>\n\n')
}


file_or_na <- function(...) {
  f <- fs::path(...)
  if (fs::file_exists(f)) f else NA_character_
}


safe_name <- function(feature_name) {
  x <- gsub("[^A-Za-z0-9_-]+", "_", as.character(feature_name))
  x <- gsub("_+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

delta_ranks <- params$rank_cols
delta_ranks[delta_ranks == "protein_seq_id"] <- "protein_annotation"

find_delta_static_plots <- function(delta_dir, ranks = delta_ranks) {
  files <- fs::path(delta_dir, paste0("significant_static_", ranks, "_forestplot.svg"))
  names(files) <- ranks
  files[fs::file_exists(files)]
}

find_delta_interactive_plots <- function(delta_dir, ranks = delta_ranks) {
  files <- fs::path(delta_dir, paste0("significant_interactive_", ranks, "_forestplot.html"))
  names(files) <- ranks
  files[fs::file_exists(files)]
}

# New delta forest plots that combine max-delta and Stouffer T ranking
find_delta_static_maxdelta_plots <- function(delta_dir, ranks = delta_ranks) {
  files <- fs::path(delta_dir, paste0("significant_static_", ranks, "_forestplot_maxdelta.svg"))
  names(files) <- ranks
  files[fs::file_exists(files)]
}

emit_delta_forest_tabs <- function(delta_dir,
                                   interactive_mode = "link",
                                   iframe_height = "720px",
                                   include_feature_plots = FALSE,
                                   max_feature_plots = 12) {

  pretty_rank <- function(x) {
    if (x == "protein_seq_id") return("Protein ID")
    if (x == "protein_annotation") return("Protein Annotation")
    stringr::str_to_title(gsub("_", " ", x))
  }

  static_files <- find_delta_static_plots(delta_dir)
  interactive_files <- find_delta_interactive_plots(delta_dir)
  maxdelta_static_files <- find_delta_static_maxdelta_plots(delta_dir)

  ranks_present <- unique(c(names(static_files), names(interactive_files), names(maxdelta_static_files)))

  if (length(ranks_present) == 0) {
    old_static <- file_or_na(delta_dir, "significant_static_forestplot.svg")
    old_inter  <- file_or_na(delta_dir, "significant_interactive_forestplot.html")

    if (!is.na(old_static) || (!is.na(old_inter) && interactive_mode != "none")) {
      ranks_present <- "all"
      if (!is.na(old_static)) static_files["all"] <- old_static
      if (!is.na(old_inter))  interactive_files["all"] <- old_inter
    }
  }

  if (length(ranks_present) == 0) {
    emit_missing("_No DELTA forest plots available._")
    return(invisible(NULL))
  }

  ranks_present <- delta_ranks[delta_ranks %in% ranks_present]

  cat_md("::: {.panel-tabset}\n\n")

  for (rk in ranks_present) {
    tab_title <- pretty_rank(rk)
    emit_heading(tab_title, level = 4)

    # Original Stouffer-based forest plot
    if (rk %in% names(static_files)) {
      emit_img(
        static_files[[rk]],
        caption = paste(tab_title, "significant forest plot sorted by Stouffer T")
      )
    } else {
      emit_missing("_Static forest plot not available._")
    }


    if (rk %in% names(interactive_files)) {
      emit_interactive_html(
        file   = interactive_files[[rk]],
        title  = paste("Interactive", tab_title, "forest plot"),
        mode   = interactive_mode,
        height = iframe_height
      )
    }

    # New max-delta/Stouffer combined forest plot
    if (rk %in% names(maxdelta_static_files)) {
      emit_img(
        maxdelta_static_files[[rk]],
        caption = paste(tab_title, "significant forest plot sorted by max delta and Stouffer T")
      )
    } else {
      emit_missing("_Max-delta/Stouffer forest plot not available for this rank._")
    }


    # Individual feature plots
    if (isTRUE(include_feature_plots)) {
      emit_heading(
        paste("Top", max_feature_plots, stringr::str_to_title(rk), "individual DELTA feature plots"),
        level = 5
      )
      emit_delta_feature_plots_for_rank(
        delta_dir = delta_dir,
        rank_name = rk,
        top_n = max_feature_plots
      )
    }

    cat_md("\n")
  }

  cat_md(":::\n\n")
}


emit_pop_scatter_tabs <- function(plots_dir,
                                  interactive_mode = "link",
                                  iframe_height = "720px") {

  pop_ranks <- c("peptide_id", params$rank_cols)

  pretty_rank <- function(x) {
    if (x == "peptide_id") return("Peptide ID")
    if (x == "protein_seq_id") return("Protein ID")
    stringr::str_to_title(gsub("_", " ", x))
  }

  static_files <- fs::path(plots_dir, paste0(pop_ranks, "_static.svg"))
  names(static_files) <- pop_ranks
  static_files <- static_files[fs::file_exists(static_files)]

  interactive_files <- fs::path(plots_dir, paste0(pop_ranks, "_interactive.html"))
  names(interactive_files) <- pop_ranks
  interactive_files <- interactive_files[fs::file_exists(interactive_files)]

  ranks_present <- union(names(static_files), names(interactive_files))

  if (length(ranks_present) == 0) {
    emit_missing("_No POP scatter plots available._")
    return(invisible(NULL))
  }

  cat_md("::: {.panel-tabset}\n\n")

  for (rk in ranks_present) {
    tab_title <- pretty_rank(rk)
    emit_heading(tab_title, level = 4)

    if (rk %in% names(static_files)) {
      emit_img(
        static_files[[rk]],
        caption = paste(tab_title, "static scatter plot")
      )
    } else {
      emit_missing("_Static plot not available._")
    }

    if (rk %in% names(interactive_files)) {
      emit_interactive_html(
        file   = interactive_files[[rk]],
        title  = paste("Interactive", tab_title, "plot"),
        mode   = interactive_mode,
        height = iframe_height
      )
    }

    cat_md("\n")
  }

  cat_md(":::\n\n")
}

emit_interactive_html <- function(file,
                                  title = "Interactive plot",
                                  mode = c("link", "embed", "none"),
                                  height = "720px") {
  mode <- match.arg(mode)

  if (length(file) == 0 || is.na(file) || !fs::file_exists(file) || mode == "none") {
    return(invisible(NULL))
  }

  rel <- safe_rel_path(file)

  cat_md(glue('<details><summary><strong>{htmlEscape(title)}</strong></summary>\n\n'))

  if (mode == "embed") {
    emit_iframe(file, height = height)
  } else if (mode == "link") {
    cat_md(glue('<p><a href="{rel}" target="_blank">Open interactive HTML</a></p>\n\n'))
  }

  cat_md('</details>\n\n')
}



get_delta_rank_table <- function(delta_dir, rank_name, top_n = 12) {

  # --------------------------------------------------------------------------
  # 1. Preferred source:
  #    new top table created only when there is at least one directional hit
  # --------------------------------------------------------------------------

  top_tbl_file <- file_or_na(
    delta_dir,
    paste0("delta_table_m_eff_", params$min_eff,"_TOP_", rank_name, "_sort_by_maxDelta-Stouffer.csv")
  )

  if (!is.na(top_tbl_file)) {
    df <- read_table_auto(top_tbl_file)

    if (!"feature" %in% names(df)) {
      return(NULL)
    }

    if (nrow(df) == 0) {
      return(df)
    }

    df <- df |>
      dplyr::filter(!is.na(.data$feature), .data$feature != "") |>
        dplyr::arrange(dplyr::desc(.data$weighted_score))

    return(dplyr::slice_head(df, n = min(top_n, nrow(df))))
  }
}


emit_delta_feature_plots_for_rank <- function(delta_dir,
                                              rank_name,
                                              top_n = 12) {
  feat_dir <- fs::path(delta_dir, "interesting_features")

  if (!fs::dir_exists(feat_dir)) {
    return(invisible(NULL))
  }

  rank_tbl <- get_delta_rank_table(
    delta_dir = delta_dir,
    rank_name = rank_name,
    top_n = top_n
  )

  if (
    is.null(rank_tbl) ||
      nrow(rank_tbl) == 0 ||
      !"feature" %in% names(rank_tbl)
  ) {
    emit_missing("_No significant DELTA features available for this rank._")
    return(invisible(NULL))
  }

  is_protein_annotation <- identical(
    rank_name,
    "protein_annotation"
  )

  # For protein_annotation tables:
  #   feature        = annotation shown in the report
  #   protein_seq_id = identifier used in filenames
  if (is_protein_annotation) {
    if (!"protein_seq_id" %in% names(rank_tbl)) {
      emit_missing(
        "_The protein annotation table does not contain `protein_seq_id`, so individual plot files cannot be matched._"
      )
      return(invisible(NULL))
    }

    rank_tbl <- rank_tbl |>
      dplyr::mutate(
        display_feature = trimws(as.character(.data$feature)),
        file_feature = trimws(as.character(.data$protein_seq_id))
      )
  } else {
    rank_tbl <- rank_tbl |>
      dplyr::mutate(
        display_feature = trimws(as.character(.data$feature)),
        file_feature = trimws(as.character(.data$feature))
      )
  }

  rank_tbl <- rank_tbl |>
    dplyr::filter(
      !is.na(.data$display_feature),
      .data$display_feature != "",
      !is.na(.data$file_feature),
      .data$file_feature != ""
    ) |>
    dplyr::distinct(.data$file_feature, .keep_all = TRUE)

  shown_idx <- 0L

  for (i in seq_len(nrow(rank_tbl))) {
    file_feature <- rank_tbl$file_feature[[i]]
    display_feature <- rank_tbl$display_feature[[i]]

    # Use protein_seq_id for matching files, e.g. protseq_112XX
    feat_safe <- safe_name(file_feature)

    delta_file <- fs::path(
      feat_dir,
      paste0(feat_safe, "_deltaplot_static.svg")
    )

    ecdf_file <- fs::path(
      feat_dir,
      paste0(feat_safe, "_ecdfplot_static.svg")
    )

    scatter_file <- fs::path(
      feat_dir,
      paste0(feat_safe, "_scatter_static.svg")
    )

    scatter_interactive_file <- fs::path(
      feat_dir,
      paste0(feat_safe, "_scatter_interactive.html")
    )

    has_delta <- fs::file_exists(delta_file)
    has_ecdf <- fs::file_exists(ecdf_file)
    has_scatter <- fs::file_exists(scatter_file)
    has_scatter_interactive <- fs::file_exists(
      scatter_interactive_file
    )

    if (!(
      has_delta ||
        has_ecdf ||
        has_scatter ||
        has_scatter_interactive
    )) {
      next
    }

    shown_idx <- shown_idx + 1L

    if (shown_idx == 1L) {
      cat_md("<details open>\n")
    } else {
      cat_md("<details>\n")
    }

    # Show annotation prominently, while retaining protein ID
    if (is_protein_annotation) {
      summary_label <- glue::glue(
        "{htmlEscape(display_feature)} ",
        "<span style=\"font-weight:normal;color:#666;\">",
        "({htmlEscape(file_feature)})",
        "</span>"
      )
    } else {
      summary_label <- htmlEscape(display_feature)
    }

    cat_md(
      glue::glue(
        "<summary><strong>{summary_label}</strong></summary>\n\n"
      )
    )

    cat_md('<div style="margin-top:0.75rem;">\n')

    # Use the annotation in the visible captions
    top_files <- c(
      delta_file,
      ecdf_file
    )[c(has_delta, has_ecdf)]

    top_labels <- c(
      paste0(display_feature, " — delta plot"),
      paste0(display_feature, " — ECDF plot")
    )[c(has_delta, has_ecdf)]

    if (length(top_files) > 0) {
      emit_img_grid(
        files = top_files,
        labels = top_labels,
        ncol = 2
      )
    }

    if (has_scatter) {
      src <- safe_rel_path(scatter_file)
      cap <- paste0(display_feature, " — scatter plot")

      cat_md('<div style="margin-top:-0.25rem;">\n')
      cat_md(
        paste0(
          '<div style="width:100%; aspect-ratio:1 / 1;',
          ' border:1px solid #e5e7eb; border-radius:8px;',
          ' overflow:hidden;">\n'
        )
      )

      cat_md(
        glue::glue(
          '<img src="{src}" ',
          'alt="{htmlEscape(cap)}" ',
          'style="width:100%; height:100%; object-fit:contain;',
          ' display:block; background:white;"/>\n'
        )
      )

      cat_md("</div>\n")

      cat_md(
        glue::glue(
          '<div style="font-size:0.92rem;color:#555;',
          'margin-top:0.35rem;">',
          "{htmlEscape(cap)}",
          "</div>\n"
        )
      )

      cat_md("</div>\n\n")
    }

    if (has_scatter_interactive) {
      rel_interactive <- safe_rel_path(
        scatter_interactive_file
      )

      cat_md(
        glue::glue(
          '<p style="margin-top:0.4rem; margin-bottom:0.2rem;">',
          '<a href="{rel_interactive}" target="_blank">',
          "Open interactive scatter HTML",
          "</a></p>\n\n"
        )
      )
    }

    cat_md("</div>\n")
    cat_md("</details>\n\n")
  }

  if (shown_idx == 0L) {
    emit_missing(
      "_No matching individual DELTA feature plot files found for this rank._"
    )
  }

  invisible(NULL)
}
```



## Analysis summary

```{r, results='asis'}
phiper_version <- as.character(utils::packageVersion("phiper"))
cat_md(glue("**PHIPER version:** `{phiper_version}`  \n"))
base_dir_display <- fs::path_rel(base_dir, start = fs::path_dir(fs::path_dir(base_dir)))
cat_md(glue("**Base results directory:** `{base_dir_display}`  \n"))
cat_md(glue("**Results folder:** `{group_col}`  \n"))
cat_md(glue("**Comparisons detected:** {length(comparisons)}\n\n"))
cat_md(paste0("- `", comparisons, "`", collapse = "\n"), "\n\n")
cat_md(
  "This report is designed to live **inside the selected results folder** so that SVG and HTML outputs can be referenced with relative paths.  \n",
  glue("POP interactive mode: `{pop_interactive_mode}`  \n"),
  glue("DELTA interactive mode: `{delta_interactive_mode}`\n\n")
)
```

## How to read this report

This report summarizes PHIPER results for a grouping variable and its pairwise comparisons.

**Enrichment counts**  
This panel shows how many peptide enrichments are detected per sample. It provides a quick overview of the overall signal level and whether some groups tend to have broader or narrower antibody reactivity profiles.

**Alpha diversity**  
Alpha diversity summarizes the diversity of enriched peptides within each sample.

- **Richness**: the number of distinct enriched features (peptides/species) observed in a sample.

**Beta diversity**  
Beta diversity summarizes differences between samples based on their peptide reactivity profiles.

- **PCoA** projects between-sample distances into a lower-dimensional space; nearby samples are more similar.
- **t-SNE** is another low-dimensional visualization that emphasizes local similarity structure.
- **Distance to centroid / dispersion plot** shows how spread out samples are within each group; larger distances indicate higher within-group heterogeneity.

When shown, **PERMANOVA** tests whether groups differ in overall composition, while the **dispersion test** evaluates whether groups differ in within-group spread.

**Longitudinal stability**  
When paired longitudinal data are available, this section summarizes how stable each subject’s peptide reactivity profile is across timepoints.

The stability analysis uses the **Kulczynski similarity** metric:

- values closer to **1** indicate more similar/stable profiles between paired timepoints
- values closer to **0** indicate larger changes over time

The combined heatmap and boxplot help identify whether some groups show more stable or more variable antibody reactivity patterns across longitudinal samples.

**POP framework**  
The POP framework compares feature prevalence between groups.

At peptide level:  
- **1** means present/enriched  
- **0** means absent/not enriched

Prevalence is the proportion of samples in a group with value 1.

For higher ranks such as species, genus, family, or order, a sample is counted as present if at least one peptide assigned to that rank is present in that sample. **For ranks above species, the prevalence comparisons are not as meaningful due to the likelihood of many samples having at least one peptide from that rank, but they are still included for completeness.** 

The POP scatterplots compare prevalence in group 1 vs group 2:  
- points near the diagonal have similar prevalence  
- points away from the diagonal differ between groups

**DELTA framework**  
The DELTA framework highlights features with the strongest shifts between groups.

Table (delta_table*csv) shows features with their standardized significance difference (Stouffer T, T_obs_stand), raw permutation p-values (p_perm), and max delta (maximum variance in prevalence). The table interesting cols are: T_obs_stand, p_perm and max_delta.

- **Forest plot**: ranked summary of features with strongest standardized differences
- **Delta plot**: peptide-level shift pattern for one selected feature
- **ECDF plot**: cumulative distribution of peptide prevalence values across the two groups
- **Scatter plot**: prevalence of the feature across all samples, colored by significance

<span style="color:#A61C1C;"><strong>POP provides an intuitive prevalence comparison, while DELTA highlights directional and standardized differences.</strong></span>

```{r, results='asis'}
for (cmp in comparisons) {
  cmp_dir <- fs::path(group_dir, cmp)
  pairwise <- cmp != "all"

  emit_heading(cmp, level = 2)

  if (!fs::dir_exists(cmp_dir)) {
    emit_missing(glue("_Comparison folder not found: {cmp_dir}_"))
    next
  }

  alpha_dir <- fs::path(cmp_dir, "alpha_diversity")
  beta_dir  <- fs::path(cmp_dir, "beta_diversity")
  stability_dir <- fs::path(cmp_dir, "similarity")
  pop_dir   <- fs::path(cmp_dir, "POP_framework")
  delta_dir <- fs::path(cmp_dir, "DELTA_framework")

  emit_heading("Overview", level = 3)
  emit_img(file_or_na(cmp_dir, "enrichment_counts.svg"), caption = "Enrichment counts")

  # Alpha diversity section
  emit_heading("Alpha diversity", level = 3)
  alpha_files <- c(
    file_or_na(alpha_dir, "plot.svg"),
    file_or_na(alpha_dir, "plot_species.svg")
  )
  alpha_labels <- c("Peptides alpha diversity plot", "Species alpha diversity plot")
  emit_img_grid(alpha_files[!is.na(alpha_files)], alpha_labels[!is.na(alpha_files)], ncol = 2)

  if (isTRUE(params$include_alpha_tables)) {
    emit_dt_details(
      file_or_na(alpha_dir, "table.xlsx"),
      "Alpha diversity table",
      top_n = params$top_n
    )
    emit_dt_details(
      file_or_na(alpha_dir, "table_species.xlsx"),
      "Alpha diversity species table",
      top_n = params$top_n
    )
  }

  # Beta diversity section
  emit_heading("Beta diversity", level = 3)
  beta_static <- c(
    file_or_na(beta_dir, "tsne2d_plot.svg"),
    file_or_na(beta_dir, "pcoa_plot.svg"),
    file_or_na(beta_dir, "dispersion_plot.svg"),
    file_or_na(beta_dir, "scree_plot.svg"),
    file_or_na(beta_dir, "capscale_plot.svg")
  )
  beta_labels <- c(
    "t-SNE 2D",
    "PCoA",
    "Dispersion plot",
    "Scree plot",
    "CAPSCALE"
  )
  emit_img_grid(beta_static[!is.na(beta_static)], beta_labels[!is.na(beta_static)], ncol = 2)

  tsne3d_html <- file_or_na(beta_dir, "tsne3d_plot.html")
  if (isTRUE(params$include_tsne3d) && !is.na(tsne3d_html)) {
    cat_md('<details open><summary><strong>Interactive t-SNE 3D</strong></summary>\n\n')
    emit_iframe(tsne3d_html, height = "720px")
    cat_md('</details>\n\n')
  }

  if (isTRUE(params$include_beta_tables)) {
    emit_dt_details(
      file_or_na(beta_dir, "permanova_results.rds"),
      "PERMANOVA results",
      top_n = params$top_n
    )
    emit_dt_details(
      file_or_na(beta_dir, "dispersion_results.rds"),
      "Dispersion results",
      top_n = params$top_n
    )
  }

  #stability, POP and DELTA frameworks are only relevant for pairwise comparisons, so we check for the existence of their directories before emitting those sections
  
  # stability section
  stability_plot <- file_or_na(stability_dir, "kulczynski_combined.svg")
  if (isTRUE(params$include_longitudinal_stability) && 
  pairwise && 
  fs::dir_exists(stability_dir) && 
  !is.na(stability_plot)) {
    emit_heading("Stability", level = 3)
    
    emit_text(
      "This section summarizes within-subject longitudinal stability using the Kulczynski similarity metric. Higher similarity indicates more stable peptide reactivity profiles between paired timepoints."
      )
      
    emit_img(
      stability_plot,
      caption = "Longitudinal stability: Kulczynski similarity heatmap and boxplot",
      width = "100%"
      )

  }

  # POP framework section
  if (pairwise && fs::dir_exists(pop_dir)) {
    emit_heading("POP framework", level = 3)

    if (isTRUE(params$include_pop_tables)) {
      emit_dt_details(
        file_or_na(pop_dir, "taxa_ranks.csv"),
        "POP taxa ranks (sorted by p_raw)",
        top_n = params$top_n
      )
      
      emit_dt_details(
        file_or_na(pop_dir, "single_peptide.csv"),
        "POP single peptide table (sorted by p_raw)",
        top_n = params$top_n
      )
    }

    plots_dir <- fs::path(pop_dir, "plots")
    if (fs::dir_exists(plots_dir) &&
        (isTRUE(params$include_pop_plots) || pop_interactive_mode != "none")) {
    emit_pop_scatter_tabs(
        plots_dir = plots_dir,
        interactive_mode = pop_interactive_mode,
        iframe_height = "720px"
    )
    }
  }


  # DELTA section
  if (pairwise && fs::dir_exists(delta_dir)) {
    emit_heading("DELTA framework", level = 3)
    emit_delta_forest_tabs(
      delta_dir = delta_dir,
      interactive_mode = delta_interactive_mode,
      iframe_height = "720px",
      include_feature_plots = isTRUE(params$include_delta_feature_plots),
      max_feature_plots = params$max_delta_feature_plots
    )

    if (isTRUE(params$include_delta_tables)) {
      emit_dt_details(
        file_or_na(delta_dir, paste0("delta_table_m_eff_gt", params$min_eff, "_p_perm_lt0.05.csv")),
        paste0("DELTA table (p_perm <= 0.05; m_eff > ", params$min_eff, "; sorted by absolute T_obs_stand, max delta and p_perm)"),
        top_n = 100
      )
    }
  }
}
```
